How is experience transformed into memory?

### 2018-08-28 paper/main.tex
checking in interim changes to intro [WIP]

What does it mean to \textit{remember} something? In traditional
episodic memory experiments \citep[e.g., list-learning or trial-based
experiments;][]{Murd62a, Kaha96}, remembering is often cast as a
discrete and binary operation: each studied item may be separated from
the rest of one's experiences, and that item may be labeled as having
been recalled versus forgotten. More nuanced studies might incorporate
self-reported confidence measures as a proxy for memory strength, or
ask participants to discriminate between ``recollecting'' the
(contextual) details of an experience or having a general feeling of
``familiarity'' \citep{Yone02}. However, characterizing and evaluating
memory in more realistic contexts (e.g., recounting a recent
experience to a friend) is fundamentally different.  First, real-world
recall is continuous, rather than discrete.  Removing an event from
the context in which it occurrs often subtantially changes its
meaning.  Second, the specific words used to describe an experience
have very little bearing on whether the experience is considered to
have been ``remembered.''  Asking whether the rememberer has precisely
reproduced a specific set of words to describe a given experience is
nearly orthogonal to whether they were actually able to remember it.
Third, one might remember the gist or essence of an experience but
forget (or neglect to recount) particular details.  Capturing the gist
of what happened is typically the ``point'' of recounting a memory to
a listener but, depending on the circumstances, accurate recall of any
specific detail may be irrelevant.

% JRM REWRITES STOPPED HERE

%(JRM: started on the next paragraph...)
Human memory studies typically assess memory for the elements within
an experience, such as the people, places and things
encountered. However, real life experiences can also be described in
terms of their temporal structure (i.e., how the contents
relate/change over time).  For instance, naturalistic experiences are
autocorrelated in space and time on short timescales. In other words,
the contents of an experience often change gradually and are therefore
similar from moment to moment. As an example, many of us spend our
mornings/evenings at home and much of our days at work (e.g., in an
office). While the specific happenings within these contexts differ,
they are often highly correlated (and thus predictable) within an
experience. Furthermore, our experiences can also be characterized by
longer timescale correlations. For example, consider returning to your
office after taking a lunch break. The spatial contexts before and
after lunch are highly similar, but separated in time. Thus, while
evaluating the ability to recall the isolated elements of an
experience has been a fruitful way to learn about human memory,
another important dimension to consider is the temporal structure of
the remembered experience, and how the temporal structure of the
memory relates to the original experience. These recurrent and
gradually changing temporal dynamics are not typically present in
traditional memory studies, but are likely critical to understanding
how our memory systems encode, store and retrieve our life
experiences.

While our experiences often change gradually, they can also sometimes change suddenly (e.g., walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While it's undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for real-life experiences. Real-life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory must not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.

>>>>

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[also see][for a review]{KoriGold94}.  First, real world recall is continuous, rather than discrete.  Removing an event from the context in which it occurrs often subtantially changes its meaning.  Second, the specific words used to describe an experience have very little bearing on whether the experience is considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  Third, one might remember the gist or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.

How might one go about formally characterizing the ``gist'' of an experience?  Real world experiences unfold over extended time intervals.  Any given moment of an experience might derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g.\ ][]{LernEtal11}.  This interplay between our subjective experiences at different timescales defines a \textit{context} for each new moment, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Further, our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated).  This allows us to form stable estimates of our current situation and thereby behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  
%%%JRM STOPPED HERE
Although our experiences often change gradually, they occasionaly change suddenly~\citep[e.g., walking through a doorway][]{RadvZack17}.

Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}. Short-timescale autocorrelations in the contents of an experience are thought to associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, EzzyDava11, DuBrDava13}.  In other words, the interplay between stable and transient temporal dynamics may impose structure on memories for those experiences. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people verbally recount their experiences in a way that is structured around events has not been established.

@@

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While it's undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for real life experiences. Real life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory must not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.
